Timeliness of security updates for Android

### 2014-10-24 security_scoring.tex
Use todolater for things we don't need to get done before submission and hence hide them from view

\subsubsection{Results: Security scores}
The scores across the whole of android are that \daMeanInsecurityPerc\ of devices are exposed to known root exploits.
There are on average \daMeanOutstandingVulnerabilities\ outstanding vulnerabilities not fixed on any device.
Only on average \daUpdatednessPerc\ of devices run the most recent version of Android.
This gives a security score of \daSecurityScore / 10.
\daTabSecScoressummary
However there are a wide variety of scores depending on the source of the device.
There have been many reports\todo{citation} that Google's Nexus devices are better at getting updates than other Android devices because Google makes the original updates and ships them to its devices.
Table~\ref{tab:sec_summary} shows that this is the case with Nexus devices getting much better scores than non-Nexus devices.
\daTabSecScoresmanufacturer
Different manufacturers have very different scores, Table~\ref{tab:sec_manufacturer} shows the scores for the \daNumSigManufacturers\ manufacturers with a significant presence in our data with \daSecScoreBestmanufacturer\ (\daSecScoreBestmanufacturerScore\ / 10) doing best and \daSecScoreWorstmanufacturer\ (\daSecScoreWorstmanufacturerScore\ / 10) doing worst.
Manufacturers are considered significant if we have data from at least \daSigNumDevices\ devices and at least \daSigNumDeviceDays\ days of contributions.
Additionally for $m$ and $u$ we ignore days with less than \daSigNumDevicesDay\ devices contributing to that day's score.

>>>>

\subsubsection{Results: Security scores}
The scores across the whole of android are that \daMeanInsecurityPerc\ of devices are exposed to known root exploits.
There are on average \daMeanOutstandingVulnerabilities\ outstanding vulnerabilities not fixed on any device.
Only on average \daUpdatednessPerc\ of devices run the most recent version of Android.
This gives a security score of \daSecurityScore / 10.
\daTabSecScoressummary
However there are a wide variety of scores depending on the source of the device.
There have been many reports\todolater{citation} that Google's Nexus devices are better at getting updates than other Android devices because Google makes the original updates and ships them to its devices.
Table~\ref{tab:sec_summary} shows that this is the case with Nexus devices getting much better scores than non-Nexus devices.
\daTabSecScoresmanufacturer
Different manufacturers have very different scores, Table~\ref{tab:sec_manufacturer} shows the scores for the \daNumSigManufacturers\ manufacturers with a significant presence in our data with \daSecScoreBestmanufacturer\ (\daSecScoreBestmanufacturerScore\ / 10) doing best and \daSecScoreWorstmanufacturer\ (\daSecScoreWorstmanufacturerScore\ / 10) doing worst.
Manufacturers are considered significant if we have data from at least \daSigNumDevices\ devices and at least \daSigNumDeviceDays\ days of contributions.
Additionally for $m$ and $u$ we ignore days with less than \daSigNumDevicesDay\ devices contributing to that day's score.

### 2014-10-24 securityupdates.tex
Use todolater for things we don't need to get done before submission and hence hide them from view

\documentclass[conference,a4paper,twoside]{IEEEtran}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[space]{grffile}
\graphicspath{figures/}
\usepackage{import}
\usepackage{url}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{todonotes}
\let\OldTodo\todo
\renewcommand{\todo}{\OldTodo[inline]}
\RequirePackage[date=terse, isbn=true, doi=true, url=false, urldate=iso8601, maxbibnames=9, backref=false, backend=bibtex]{biblatex}
\addbibresource{securityupdates.bib}
\renewcommand{\bibfont}{\small}

@@

\section{Introduction}
Android has \percMarketShare\ of the smartphone market and while the core development is controlled by Google there are about \daNumManufacturers\footnote{\label{foot:dadata}We computed this from the \da~\cite{Wagner2013} data, see \S\ref{sec:android_update_process}.} manufacturers which make devices which run Android.
Android receives regular updates  which add new features and fix bugs, including security vulnerabilities.
Many of these manufacturers customise the version of Android they ship and sometimes network operators (of which there are at least \daNumOperators\dafoot) make further modifications.
Hence when Google produces an update to Android, the update may have to pass through the manufacturer and operator before reaching the user.
The manufacturer and operator have little financial incentive to perform this work as they have already sold the device.
Operators are cautious because if they ship a broken update then they will get many support requests which is expensive.
However there is ongoing legal action to force operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.\todo{Check on the status of this legal action}
Many phones are sold on long (12--24 month) contracts with monthly payments yet stop receiving updates before the end of the contract,\todo{Is there a survey of the distribution of contract lengths and phone purchase methods?} while Windows XP could be purchased for a one off payment in October 2001 and received security updates until April 2014.
Corporate or public sector buyers are encouraged to purchase secure devices, but we have found little concrete guidance. For example, CESG, which advises the UK government on how to secure its computer systems, recommends picking Android devices from manufactures which are good at shipping security updates promptly~\cite{CESG2013} but it does not state which manufacturers are better.\todo{Which other advisory organisations give that advice?}
We provide this information in~\S\ref{sec:security_scoring}.

@@

\subsection{Versions of Android running on devices}
We need historical data on the distributions of versions of Android running on devices and additional data about those devices to allows us to investigate variation between devices and confirm the reliability of these data.
The only suitable data which is available for this is the \da\ data~\cite{Wagner2013}.
The \da\ data contains data from \daNumDevices\ devices with a total of \daDeviceDays\footnote{Here we are only counting devices and days for which we have valid OS version data} device days, it contains data for more than \daMonths\ months for \daMonthsDevices\ devices.
\todo{ talk about the \da\ project a bit}
Various different kinds of data\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/collected.htm}} are collected by the \da\ Android app\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/}}.
Among these are the build string and API version for the version of Android currently running on the device each day.
The API version is a well defined integer, however it does not always change with new Android releases, particularly security bug fixes should not increase this value.
The build string is `The user-visible version string', fortunately most (99.9\%)\todo{use a generated number} devices in the data have a build string of the form `x.y.z random string' and so it is possible to extract the Android version number.
On a large proportion of devices `random string' is a well defined build number\footnote{\url{https://source.android.com/source/build-numbers.html}} but different manufacturers use different schema.

@@

To compute the latency between upstream releases and their inclusion in Android we scraped the release pages for those projects, to obtain the version numbers and release dates.
This allows us to compute the latency between an upstream project being released and it being included in Android, this is shown in Table~\ref{tab:update_ecosystem}.
\begin{table}
\centering
\begin{tabular}{l|r|r}
Project	&	\# releases	&	latency \\ \hline
linux	&	\linuxNumVersions	&	\linuxMeanUpdateLatency \\
openssl	&	\opensslNumVersions	&	\opensslMeanUpdateLatency \\
bouncycastle	&	\bouncycastleNumVersions	&	\bouncycastleMeanUpdateLatency \\
\end{tabular}
\caption{Flow of updates from upstream projects into Android. Number of updates as in Figure~\ref{fig:update_ecosystem}, latency in days for all pairs of versions we have data on.\todo{scrape the other 26 websites... is it worth it?}}
\label{tab:update_ecosystem}
\end{table}

@@

Finding vulnerabilities is frequently a time consuming manual process but, Brumley et al.~\cite{Brumley2008} showed that it is possible to automatically generate exploits from binary patches.
In a similar way some of the vulnerabilities which have been exploited were publicly found by examining the commit made to the Android which fixed the vulnerability.
This effect has been observed before~\cite{Barth2011} in the Firefox source code repositories, Google does not release the source code until after the release of the update reducing this effect to the delay between an update being released and it reaching the last end user device.
There have been other attempts to automatically find vulnerabilities.
The Woodpecker tool~\cite{Grace2012} automatically finds permission leaks in stock Android phone images.
The update process itself can allow Android apps to gain privilege through `Pileup' vulnerabilities~\cite{Xing2014} by registering for new permissions before the update which creates that permission is installed.

@@

Examining security updates has been a key way of assessing security for many years.
Many software update systems have been found not to authenticate the connection to the update server and/or do not authenticate the downloaded binaries~\cite{Bellissimo2006}.
Even package management systems designed to provide secure updates have been found to have vulnerabilities~\cite{Cappos2008}.
Fortunately Android does authenticate the update binaries (though the APK vulnerabilities circumvented this) and Google Play downloads them over a secure connection~\cite{Viennot2014}. \todo{double check that}

The updatedness rate for Android of \daUpdatednessPerc\ compares unfavourable with the more than 90\% rate for Windows XP SP2 computers contacting the Microsoft update servers~\cite{Gkantsidis2006} though that is within one OS version and only for those computers which did connect to the servers -- but that is the default.
However 27\% of Windows computers were still running Windows XP in July 2014\footnote{\url{https://archive.today/PLGxn}} -- four months after it went out of security support.

>>>>

\documentclass[conference,a4paper,twoside]{IEEEtran}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[space]{grffile}
\graphicspath{figures/}
\usepackage{import}
\usepackage{url}
\usepackage{hyperref}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{todonotes}
\let\OldTodo\todo
\renewcommand{\todo}{\OldTodo[inline]}
\newcommand{\todolater}{}%\todo
\RequirePackage[date=terse, isbn=true, doi=true, url=false, urldate=iso8601, maxbibnames=9, backref=false, backend=bibtex]{biblatex}
\addbibresource{securityupdates.bib}
\renewcommand{\bibfont}{\small}

@@

\section{Introduction}
Android has \percMarketShare\ of the smartphone market and while the core development is controlled by Google there are about \daNumManufacturers\footnote{\label{foot:dadata}We computed this from the \da~\cite{Wagner2013} data, see \S\ref{sec:android_update_process}.} manufacturers which make devices which run Android.
Android receives regular updates  which add new features and fix bugs, including security vulnerabilities.
Many of these manufacturers customise the version of Android they ship and sometimes network operators (of which there are at least \daNumOperators\dafoot) make further modifications.
Hence when Google produces an update to Android, the update may have to pass through the manufacturer and operator before reaching the user.
The manufacturer and operator have little financial incentive to perform this work as they have already sold the device.
Operators are cautious because if they ship a broken update then they will get many support requests which is expensive.
However there is ongoing legal action to force operators to ship updates for security vulnerabilities~\cite{Soghoian2013}.\todolater{Check on the status of this legal action}
Many phones are sold on long (12--24 month) contracts with monthly payments yet stop receiving updates before the end of the contract,\todolater{Is there a survey of the distribution of contract lengths and phone purchase methods?} while Windows XP could be purchased for a one off payment in October 2001 and received security updates until April 2014.
Corporate or public sector buyers are encouraged to purchase secure devices, but we have found little concrete guidance. For example, CESG, which advises the UK government on how to secure its computer systems, recommends picking Android devices from manufactures which are good at shipping security updates promptly~\cite{CESG2013} but it does not state which manufacturers are better.\todolater{Which other advisory organisations give that advice?}
We provide this information in~\S\ref{sec:security_scoring}.

@@

\subsection{Versions of Android running on devices}
We need historical data on the distributions of versions of Android running on devices and additional data about those devices to allows us to investigate variation between devices and confirm the reliability of these data.
The only suitable data which is available for this is the \da\ data~\cite{Wagner2013}.
The \da\ data contains data from \daNumDevices\ devices with a total of \daDeviceDays\footnote{Here we are only counting devices and days for which we have valid OS version data} device days, it contains data for more than \daMonths\ months for \daMonthsDevices\ devices.
\todo{ talk about the \da\ project a bit}
Various different kinds of data\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/collected.htm}} are collected by the \da\ Android app\footnote{\url{https://deviceanalyzer.cl.cam.ac.uk/}}.
Among these are the build string and API version for the version of Android currently running on the device each day.
The API version is a well defined integer, however it does not always change with new Android releases, particularly security bug fixes should not increase this value.
The build string is `The user-visible version string', fortunately most (99.9\%)\todolater{use a generated number} devices in the data have a build string of the form `x.y.z random string' and so it is possible to extract the Android version number.
On a large proportion of devices `random string' is a well defined build number\footnote{\url{https://source.android.com/source/build-numbers.html}} but different manufacturers use different schema.

@@

To compute the latency between upstream releases and their inclusion in Android we scraped the release pages for those projects, to obtain the version numbers and release dates.
This allows us to compute the latency between an upstream project being released and it being included in Android, this is shown in Table~\ref{tab:update_ecosystem}.
\begin{table}
\centering
\begin{tabular}{l|r|r}
Project	&	\# releases	&	latency \\ \hline
linux	&	\linuxNumVersions	&	\linuxMeanUpdateLatency \\
openssl	&	\opensslNumVersions	&	\opensslMeanUpdateLatency \\
bouncycastle	&	\bouncycastleNumVersions	&	\bouncycastleMeanUpdateLatency \\
\end{tabular}
\caption{Flow of updates from upstream projects into Android. Number of updates as in Figure~\ref{fig:update_ecosystem}, latency in days for all pairs of versions we have data on.\todolater{scrape the other 26 websites... is it worth it?}}
\label{tab:update_ecosystem}
\end{table}

@@

Finding vulnerabilities is frequently a time consuming manual process but, Brumley et al.~\cite{Brumley2008} showed that it is possible to automatically generate exploits from binary patches.
In a similar way some of the vulnerabilities\todolater{which?} which have been exploited were publicly found by examining the commit made to the Android which fixed the vulnerability.
This effect has been observed before~\cite{Barth2011} in the Firefox source code repositories, Google does not release the source code until after the release of the update reducing this effect to the delay between an update being released and it reaching the last end user device.
There have been other attempts to automatically find vulnerabilities.
The Woodpecker tool~\cite{Grace2012} automatically finds permission leaks in stock Android phone images.
The update process itself can allow Android apps to gain privilege through `Pileup' vulnerabilities~\cite{Xing2014} by registering for new permissions before the update which creates that permission is installed.

@@

Examining security updates has been a key way of assessing security for many years.
Many software update systems have been found not to authenticate the connection to the update server and/or do not authenticate the downloaded binaries~\cite{Bellissimo2006}.
Even package management systems designed to provide secure updates have been found to have vulnerabilities~\cite{Cappos2008}.
Fortunately Android does authenticate the update binaries (though the APK vulnerabilities circumvented this) and Google Play downloads them over a secure connection~\cite{Viennot2014}. \todolater{double check that}

The updatedness rate for Android of \daUpdatednessPerc\ compares unfavourable with the more than 90\% rate for Windows XP SP2 computers contacting the Microsoft update servers~\cite{Gkantsidis2006} though that is within one OS version and only for those computers which did connect to the servers -- but that is the default.
However 27\% of Windows computers were still running Windows XP in July 2014\footnote{\url{https://archive.today/PLGxn}\todolater{Get updated figures before publication}} -- four months after it went out of security support.
